Contracts for unions and intersections are hard for Nickel

### 2021-05-25 paper.tex
[WIP] Add intro, rework typing

>>>>

Build systems, package managers, operating systems, cloud infrastructure,
continuous integration and web services\resolved{Also continuous integration}
are examples of modern complex software systems which require an extensive and
non-trivial configuration. To handle this complexity, the declarative approach
has become more and more prominent: instead of applying a sequence of
transformations to a system in order to reach a desired state, this state is
described as static textual data. Some tool is then in charge of applying the
necessary transformations. This approach makes the configuration persistent,
reproducible, analyzable and simulable. Text-based configurations are led to
play an important role in critical aspects of software engineering, including
security, availability, and maintainability.

However, static text-based configurations show their limits beyond a given size
and complexity. Data cannot be transformed, combined, shared, nor validated.
This lack of expressivity results at best in boilerplate and duplication of
information, but it can also make data invalid or inconsistent. Correctly
configuring a modern system is hard and failures may have substantial negative
consequences.

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems and cloud deployment
tools (Kubernetes, Terraform or NixOps) as well.

\subsection*{To type or not to type}

The end goal being data generation, Nickel is a dynamic, pure and functional
language. Our goal is that writing a simple configuration in Nickel feels as
easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
configurations by being programmable, and by providing means of ensuring
correctness.

In this context, correctness means two things:
\begin{enumerate}
    \item The program runs without errors. For example, an error is raised when
        evaluating the nonsensical division \nickel{1 / "ftp"}.
    \item The program generates valid data, i.e. data that conforms to a schema
        (imposed by the system consuming the configuration). For example, when
        configuring a systemd service, the schema may require the presence of a
        field \lstinline+port+ that holds a valid port number.
\end{enumerate}

A traditional answer is to use types, and in particular static types for (1).
However, Nickel programs are of a special kind: they are usually simple,
terminating programs that run on fixed inputs. Errors of type (1) will show up
at evaluation anyway, or will not matter (dead code). For validation (2),
traditional types are not adapated: statically checking common requirements
like being a valid port number or a valid URL is, while not technically
impossible, really hard to do.

In a dynamic language, it is natural and easier to validate the data at runtime,
\textit{a posteriori}. This is the principle of dynamic typing. But dynamically
typed languages usually only check data when used in a primitive operation, like
addition. In Nickel, the interpreter could not have a clue that a field
\nickel{port}

In the configuration setting, there is a
singular dilemma with respect to typing: since a program is run on fixed inputs
and is expected to terminate, any relevant type error will show up at
evaluation.  Why bother with the complexity of a static type system? On the
other hand, as the configurated systems become more and more complex, so do
configuration codebases. Medium-sized to big codebases necessarily , static
types become attractive again.  For reusable code — that is, library functions
—, static types are specifically adapted, and bring all the usual benefits of
early error detection, code robustness, better code structure, documentation,
and so on.
